---
title: "Explaining migration peaks"
author: "Bart Hoekstra"
format:
  html:
    embed-resources: true
execute:
  cache: false
  warning: false
  message: false
---

## Setup

This workflow renders the figures from the committed SHAP results for Herwijnen (`nlhrw`), 2017–2024. Run from the repository root; see the [README](README.md) for software and data requirements. Model training is kept separately in `R/prepare_shap.R` and is never run here.

```{r}
library(data.table)
library(tidyverse)
library(mlr3verse)
library(mlr3db)
library(patchwork)
library(shapviz)
library(khroma)
library(ggh4x)
library(ggside)
library(tidytext)
library(ggstats)

source("R/functions_training.R")
years <- 2017:2024
spring_files <- sprintf("data/processed/sv_spring_%d.RDS", years)
autumn_files <- sprintf("data/processed/sv_autumn_%d.RDS", years)
missing_files <- c(spring_files, autumn_files)[!file.exists(c(spring_files, autumn_files))]
if (length(missing_files)) {
  stop("Required processed data are missing: ", paste(missing_files, collapse = ", "),
       ". Run git lfs pull from the repository root to retrieve the committed data files.")
}
dir.create("plots", showWarnings = FALSE)
```

## Nightly migration traffic

Hourly migration traffic rates are counted once per timestamp, summed by night, and ranked within each year and season. Ties receive the same minimum rank.

```{r}
spring_task <- create_task("nlhrw", "spring")
autumn_task <- create_task("nlhrw", "autumn")

spring_data <- spring_task$data(cols = spring_task$col_info$id)
autumn_data <- autumn_task$data(cols = autumn_task$col_info$id)

spring_ranks_and_datetimes <- select(spring_data, datetime, period_rank_mtr, mtr) %>%
  mutate(year = lubridate::year(datetime))
spring_nightly_mt <- spring_ranks_and_datetimes %>%
  distinct(datetime, .keep_all = TRUE) %>%
  group_by(year, period_rank_mtr) %>%
  summarise(nightly_mt = sum(mtr)) %>%
  ungroup() %>%
  group_by(year) %>%
  mutate(nightly_mt_rank = rank(-nightly_mt, ties.method = "min"))
spring_ranks_and_datetimes <- left_join(spring_ranks_and_datetimes, spring_nightly_mt)

autumn_ranks_and_datetimes <- select(autumn_data, datetime, period_rank_mtr, mtr) %>%
  mutate(year = lubridate::year(datetime))
autumn_nightly_mt <- autumn_ranks_and_datetimes %>%
  distinct(datetime, .keep_all = TRUE) %>%
  group_by(year, period_rank_mtr) %>%
  summarise(nightly_mt = sum(mtr)) %>%
  ungroup() %>%
  group_by(year) %>%
  mutate(nightly_mt_rank = rank(-nightly_mt, ties.method = "min"))
autumn_ranks_and_datetimes <- left_join(autumn_ranks_and_datetimes, autumn_nightly_mt)

cbrt_trans <- scales::trans_new(
  name = "cbrt",
  transform = function(x) sign(x) * abs(x)^(1/3),
  inverse   = function(x) sign(x) * abs(x)^3
)
```

## Stability across training years

Each SHAP file omits one year during training but explains all input rows. These panels compare mean absolute SHAP across the eight fitted models; height is omitted from display.

```{r}
sv_spring_years <- lapply(spring_files, readRDS)

sv_autumn_years <- lapply(autumn_files, readRDS)

validate_shap_rows <- function(sv_list, data) {
  for (sv in sv_list) {
    stopifnot(nrow(sv$S) == nrow(data), nrow(sv$X) == nrow(data))
    for (variable in names(sv$X)) {
      if (!isTRUE(all.equal(sv$X[[variable]], data[[variable]], check.attributes = FALSE))) {
        stop("Processed SHAP rows do not match model input rows for ", variable)
      }
    }
  }
}
validate_shap_rows(sv_spring_years, spring_data)
validate_shap_rows(sv_autumn_years, autumn_data)

sv_list_to_dataframe <- function(sv_list, season, years) {
  sv_importances <- lapply(sv_list, sv_importance, kind = "no")
  sv_importances_df <- lapply(seq_along(sv_list), function(i) {
    df <- data.frame(mean_abs_shap = sv_importances[[i]]) %>% rownames_to_column(var = "variable")
    df$year_held_out <- years[[i]]
    df
  })
  sv_importances_df <- bind_rows(sv_importances_df)
  sv_importances_df$season <- season
  sv_importances_df
}

sv_spring <- sv_list_to_dataframe(sv_spring_years, "spring", 2017:2024)
sv_autumn <- sv_list_to_dataframe(sv_autumn_years, "autumn", 2017:2024)

sv_spring %>%
  filter(variable != "height") %>%
  ggplot(aes(x = mean_abs_shap, y = reorder(variable, mean_abs_shap, FUN = mean, decreasing = FALSE), color = year_held_out)) +
  geom_point() +
  ggstats::geom_stripped_rows(linewidth = 0) +
  scale_color_continuous(name = "Year held out") +
  theme_bw(base_size = 12) +
  labs(x = "Mean absolute SHAP value", y = "Variable", title = "Spring") -> spring_shap_stability

sv_autumn %>%
  filter(variable != "height") %>%
  ggplot(aes(x = mean_abs_shap, y = reorder(variable, mean_abs_shap, FUN = mean, decreasing = FALSE), color = year_held_out)) +
  geom_point() +
  ggstats::geom_stripped_rows(linewidth = 0) +
  scale_color_continuous(name = "Year held out") +
  theme_bw(base_size = 12) +
  labs(x = "Mean absolute SHAP value", y = "Variable", title = "Autumn") -> autumn_shap_stability

ggsave(spring_shap_stability, filename = "plots/shap_stability_spring.pdf", height = 29.7/2, width = 21.0/2)
ggsave(spring_shap_stability, filename = "plots/shap_stability_spring.png", height = 29.7/2, width = 21.0/2)

ggsave(autumn_shap_stability, filename = "plots/shap_stability_autumn.pdf", height = 29.7/2, width = 21.0/2)
ggsave(autumn_shap_stability, filename = "plots/shap_stability_autumn.png", height = 29.7/2, width = 21.0/2)
```

The combined stability figure places the spring panel above the autumn panel.

```{r}
combined_stability <-
  (spring_shap_stability + guides(color = "none")) /
  (autumn_shap_stability + theme(
    legend.position = "inside",
    legend.position.inside = c(0.9, 0.5),
    legend.justification = c("right", "center"),
    legend.background = element_rect(fill = fill_alpha("white", 0.9)),
    legend.key.size = unit(2, "line"),
    legend.text = element_text(size = unit(14, "line"))
  )) + plot_layout(axes = "collect_x", axis_titles = "collect")
ggsave("plots/shap_stability.pdf", plot = combined_stability, height = 29.7/2, width = 21.0/2)
ggsave("plots/shap_stability.png", plot = combined_stability, height = 29.7/2, width = 21.0/2)
```

## Predictor importance on peak and low-traffic nights

Peak nights have rank 1–10; low-traffic nights have rank greater than 111.

```{r}
overall_spring <- sv_importance(sv_spring_years[[1]], max_display = 11, kind = "no")
top_nights_spring <- sv_importance(sv_spring_years[[1]][spring_ranks_and_datetimes$nightly_mt_rank < 11, ], kind = "no")
worst_nights_spring <- sv_importance(sv_spring_years[[1]][spring_ranks_and_datetimes$nightly_mt_rank > 111, ], kind = "no")

overall_spring <- data.frame(shap = overall_spring) %>% rownames_to_column(var = "variable") %>% mutate(level = "overall")
top_nights_spring <- data.frame(shap = top_nights_spring) %>% rownames_to_column(var = "variable") %>% mutate(level = "top")
worst_nights_spring <- data.frame(shap = worst_nights_spring) %>% rownames_to_column(var = "variable") %>% mutate(level = "worst")

overall_autumn <- sv_importance(sv_autumn_years[[1]], max_display = 11, kind = "no")
top_nights_autumn <- sv_importance(sv_autumn_years[[1]][autumn_ranks_and_datetimes$nightly_mt_rank < 11, ], kind = "no")
worst_nights_autumn <- sv_importance(sv_autumn_years[[1]][autumn_ranks_and_datetimes$nightly_mt_rank > 111, ], kind = "no")

overall_autumn <- data.frame(shap = overall_autumn) %>% rownames_to_column(var = "variable") %>% mutate(level = "overall")
top_nights_autumn <- data.frame(shap = top_nights_autumn) %>% rownames_to_column(var = "variable") %>% mutate(level = "top")
worst_nights_autumn <- data.frame(shap = worst_nights_autumn) %>% rownames_to_column(var = "variable") %>% mutate(level = "worst")
df <- bind_rows(
  bind_rows(overall_spring, top_nights_spring, worst_nights_spring) %>%
    mutate(season = "Spring"),
  bind_rows(overall_autumn, top_nights_autumn, worst_nights_autumn) %>%
    mutate(season = "Autumn")) %>%
  mutate(season = factor(season, levels = c("Spring", "Autumn"))) %>%
  filter(variable != "height") %>%
  group_by(season, level) %>%
  slice_max(order_by = shap, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    level = case_when(
      level == "overall" ~ "All",
      level == "top" ~ "Peak nights",
      level == "worst" ~ "Lowest nights",
    ),
    level = factor(level, levels = c("All", "Peak nights", "Lowest nights"))
  )

df_ord <- df %>%
  group_by(season, variable) %>%
  summarise(mean_abs_shap = mean(abs(shap), na.rm = TRUE), .groups = "drop")

df %>%
  left_join(df_ord, by = c("season", "variable")) %>%
  mutate(
    variable = case_when(
      variable == "yday" ~ "Day of year",
      variable == "rdr_t" ~ "Temperature (R)",
      variable == "solar_elev" ~ "Solar elevation",
      variable == "rdr_v" ~ "V-comp. of wind (R)",
      variable == "fl_inst_blh" ~ "Boundary layer height (E)",
      variable == "rdr_r_poly1" ~ "Humidity (R)",
      variable == "fl_inst_lsrr" ~ "Precipitation (E)",
      variable == "solar_elevation_change" ~ "Solar elev. change",
      variable == "stp_t2m_168hrs_sum" ~ "Temperature (S)",
      variable == "rdr_u" ~ "U-comp. of wind (R)",
      variable == "stp_t2m_3hrs_min" ~ "Temperature (S)",
      variable == "stp_lsrr_72hrs_min" ~ "Precipitation (S)",
      variable == "solar_azim" ~ "Solar azimuth",
      variable == "stp_mcc_168hrs_sum" ~ "Cloud cover (S)",
      variable == "stp_t2m_72hrs_max" ~ "Temperature (S)",
      variable == "nr_birds" ~ "Simulated migrants (R)",
      variable == "rdr_blh" ~ "Boundary layer height (R)",
      .default = variable),
  ) %>%
  ggplot(aes(
    x = shap,
    y = reorder_within(variable, mean_abs_shap, season),
    fill = level
  )) +
  geom_col(position = position_dodge2(reverse = TRUE, preserve = "single")) +
  facet_wrap(~ season, scales = "free") +
  labs(y = "Variable") +
  scale_y_reordered() +
  scale_fill_manual(values = c("#4269D0FF", "#3CA951FF", "#FF725CFF"), name = "") +
  labs(x = "Mean absolute SHAP") +
  theme_bw(base_size = 14) +
    theme(
    legend.position = "inside",
    legend.position.inside = c(0.35, 0.2),
    legend.justification = c("right", "center"),
    legend.background = element_blank(),
    legend.key.size = unit(2, "line"),
    legend.text = element_text(size = unit(14, "line"))
  )

ggsave("plots/predictor_ranks.pdf", width = 12, height = 6)
ggsave("plots/predictor_ranks.png", width = 12, height = 6)
```

## Wind effects

The wind surface sums the two wind-component SHAP values. Marginal densities compare background conditions with bird-density weighting; wind direction is not the same as migration direction.

```{r}
variables_of_interest <- c("rdr_u", "rdr_v", "rdr_blh", "fl_inst_lsrr", "stp_t2m_168hrs_sum")

spring <- sv_spring_years[[1]]
autumn <- sv_autumn_years[[1]]

feature_values_spring <- spring$X[variables_of_interest]
feature_values_autumn <- autumn$X[variables_of_interest]
shap_values_spring <- data.frame(spring$S[, variables_of_interest])
shap_values_autumn <- data.frame(autumn$S[, variables_of_interest])

shap_spring <- data.frame(feature_values_spring, shap = shap_values_spring) %>%
  bind_cols(spring_ranks_and_datetimes) %>%
  bind_cols(spring_data %>% select(height, dens)) %>%
  mutate(season = "Spring")

shap_autumn <- data.frame(feature_values_autumn, shap = shap_values_autumn) %>%
  bind_cols(autumn_ranks_and_datetimes) %>%
  bind_cols(autumn_data %>% select(height, dens)) %>%
  mutate(season = "Autumn")

shap <- bind_rows(shap_spring, shap_autumn) %>%
  mutate(season = factor(season, levels = c("Spring", "Autumn")))

shap_wind <- shap %>%
  mutate(shap_wind = shap.rdr_u + shap.rdr_v)

ggplot() +
  stat_summary_2d(aes(x = rdr_u, y = rdr_v, z = shap_wind), data = shap_wind, bins = 40) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  ggside::geom_xsidedensity(aes(x = rdr_u, weight = dens, y = after_stat(scaled), linetype = "Migration-weighted"),
                            data = shap_wind, key_glyph = "path") +
  ggside::geom_xsidedensity(aes(x = rdr_u, y = after_stat(scaled), linetype = "Background"),
                            data = shap_wind, key_glyph = "path") +
  ggside::geom_ysidedensity(aes(y = rdr_v, weight = dens, x = after_stat(scaled), linetype = "Migration-weighted"),
                            data = shap_wind, key_glyph = "path") +
  ggside::geom_ysidedensity(aes(y = rdr_v, x = after_stat(scaled), linetype = "Background"),
                            data = shap_wind, key_glyph = "path") +
  scale_fill_vik(name = "SHAP ") +
  scale_color_viridis_c(trans = cbrt_trans) +
  scale_x_continuous(breaks = seq(from = -30, to = 50, by = 10)) +
  scale_y_continuous(breaks = seq(from = -30, to = 40, by = 10)) +
  scale_xsidey_continuous(breaks = c(0, 1)) +
  scale_ysidex_continuous(guide = guide_axis(angle = -90), breaks = c(0, 1)) +
  scale_linetype_manual(
    name   = "Density type",
    values = c("Background" = "dashed",
               "Migration-weighted" = "solid"),
    breaks = c("Background", "Migration-weighted"),
    drop   = FALSE
  ) +
  coord_fixed() +
  facet_wrap(vars(season)) +
  labs(x = "U component of wind [m/s]", y = "V component of wind [m/s]") +
  theme_bw(base_size = 14) +
  theme(ggside.panel.scale = 0.1) -> shap_wind

ggsave(plot = shap_wind, filename = "plots/shap_wind.png", width = 12, height = 5)
ggsave(plot = shap_wind, filename = "plots/shap_wind.pdf", width = 12, height = 5)
```

## Other environmental effects

Lines show binned mean SHAP and ribbons the interquartile range. Seasonal tails are trimmed at the 95th percentile (also the 5th for temperature); marginal densities use nightly migration traffic as weights.

```{r}
trim_by_season <- function(df, x, p_lo = 0, p_hi = 1) {
  x <- rlang::enquo(x)

  lims <- df %>%
    group_by(season) %>%
    summarise(
      x_lo = quantile(!!x, p_lo, na.rm = TRUE),
      x_hi = quantile(!!x, p_hi, na.rm = TRUE),
      .groups = "drop"
    )

  df_trim <- df %>%
    left_join(lims, by = "season") %>%
    filter((!!x) >= x_lo, (!!x) <= x_hi) %>%
    select(-x_lo, -x_hi)

  list(data = df_trim, lims = lims)
}

ylim_effectplots <- c(-0.23, 0.23)
season_colors <- c("Autumn" = "#EFB118FF", "Spring" = "#3CA951FF")

plot_shared_data <- list(
  geom_ribbon(aes(ymin = shap_q25, ymax = shap_q75, fill = season), alpha = 0.4),
  geom_line(aes(y = shap_mean, color = season), linewidth = 0.8),
  geom_hline(yintercept = 0, color = "black", linetype = "dashed")
)

plot_shared_decorations <- list(
  scale_xsidey_continuous(breaks = c(0, 1), limits = c(0, 1.1)),
  scale_color_manual(values = season_colors, name = "Season"),
  scale_fill_manual(values = season_colors, name = "Season"),
  scale_linetype_manual(
    name   = "Density type",
    values = c("Background" = "dashed",
               "Migration-weighted" = "solid"),
    breaks = c("Background", "Migration-weighted"),
    drop   = FALSE
  ),
  theme_bw(base_size = 14),
  theme(ggside.panel.scale = 0.15),
  coord_cartesian(ylim = ylim_effectplots, expand = FALSE),
  facet_wrap(vars(location))
)

blh_trim <- trim_by_season(shap, rdr_blh, p_lo = 0, p_hi = 0.95)
shap_blh_trim <- blh_trim$data

shap_binned_blh <- shap_blh_trim %>%
  mutate(blh_bin = cut_interval(rdr_blh, n = 50)) %>%
  group_by(season, blh_bin) %>%
  summarise(
    rdr_blh_mid = mean(rdr_blh, na.rm = TRUE),
    shap_mean   = mean(shap.rdr_blh, na.rm = TRUE),
    shap_q25    = quantile(shap.rdr_blh, 0.25, na.rm = TRUE),
    shap_q75    = quantile(shap.rdr_blh, 0.75, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(location = "Radar-level")

pl_blh <- ggplot(shap_binned_blh, aes(x = rdr_blh_mid)) +
  plot_shared_data +
  ggside::geom_xsidedensity(aes(x = rdr_blh, y = after_stat(scaled),
                                weight = nightly_mt, color = season, linetype = "Migration-weighted"),
                            data = shap_blh_trim, key_glyph = "path") +
  ggside::geom_xsidedensity(aes(x = rdr_blh, y = after_stat(scaled),
                                color = season, linetype = "Background"), data = shap_blh_trim, key_glyph = "path") +
  labs(x = "Boundary layer height [m]", y = "Mean SHAP value") +
  scale_x_continuous(breaks = c(0, 250, 500, 750, 1000)) +
  plot_shared_decorations

shap_precip <- shap %>%
  mutate(fl_inst_lsrr_mmhr = fl_inst_lsrr * 3600)

prec_trim <- trim_by_season(shap_precip, fl_inst_lsrr_mmhr, p_lo = 0, p_hi = 0.95)
shap_precip_trim <- prec_trim$data

shap_binned_fl_inst_lsrr <- shap_precip_trim %>%
  mutate(lsrr_bin = cut_interval(fl_inst_lsrr_mmhr, n = 50)) %>%
  group_by(season, lsrr_bin) %>%
  summarise(
    fl_inst_lsrr_mid = mean(fl_inst_lsrr_mmhr, na.rm = TRUE),
    shap_mean = mean(shap.fl_inst_lsrr, na.rm = TRUE),
    shap_q25  = quantile(shap.fl_inst_lsrr, 0.25, na.rm = TRUE),
    shap_q75  = quantile(shap.fl_inst_lsrr, 0.75, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(location = "En-route")

pl_precip <- ggplot(shap_binned_fl_inst_lsrr, aes(x = fl_inst_lsrr_mid)) +
  plot_shared_data +
  ggside::geom_xsidedensity(aes(x = fl_inst_lsrr_mmhr, y = after_stat(scaled),
                                weight = nightly_mt, color = season, linetype = "Migration-weighted"),
                            data = shap_precip_trim, key_glyph = "path") +
  ggside::geom_xsidedensity(aes(x = fl_inst_lsrr_mmhr, y = after_stat(scaled), color = season, linetype = "Background"),
                            data = shap_precip_trim, key_glyph = "path") +
  labs(x = "Precipitation [mm/hr]", y = "Mean SHAP value") +
  plot_shared_decorations

shap_temp <- shap %>%
  mutate(stp_t2m_168hrs_avg_c = stp_t2m_168hrs_sum / 168 - 273.15)

temp_trim <- trim_by_season(shap_temp, stp_t2m_168hrs_avg_c, p_lo = 0.05, p_hi = 0.95)
shap_temp_trim <- temp_trim$data

shap_binned_temp <- shap_temp_trim %>%
  mutate(temp_bin = cut_interval(stp_t2m_168hrs_avg_c, n = 50)) %>%
  group_by(season, temp_bin) %>%
  summarise(
    temp_mid = mean(stp_t2m_168hrs_avg_c, na.rm = TRUE),
    shap_mean = mean(shap.stp_t2m_168hrs_sum, na.rm = TRUE),
    shap_q25  = quantile(shap.stp_t2m_168hrs_sum, 0.25, na.rm = TRUE),
    shap_q75  = quantile(shap.stp_t2m_168hrs_sum, 0.75, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(location = "Stopover")

pl_temp <- ggplot(shap_binned_temp, aes(x = temp_mid)) +
  plot_shared_data +
  ggside::geom_xsidedensity(aes(x = stp_t2m_168hrs_avg_c, y = after_stat(scaled),
                                weight = nightly_mt, color = season, linetype = "Migration-weighted"),
                            data = shap_temp_trim, key_glyph = "path") +
  ggside::geom_xsidedensity(aes(x = stp_t2m_168hrs_avg_c, y = after_stat(scaled),
                                color = season, linetype = "Background"), data = shap_temp_trim, key_glyph = "path") +
  labs(x = "7-day Average temperature [°C]", y = "Mean SHAP value") +
  plot_shared_decorations

pl_blh +  pl_precip + pl_temp + plot_layout(axes = "collect", guides = "collect")
ggsave("plots/shap_nonwind.png", width = 12, height = 5)
ggsave("plots/shap_nonwind.pdf", width = 12, height = 5)
```

## Group contributions

Predictor SHAP values are summed by environmental variable and location. The following summaries average across all eight fitted models, including models that trained on the displayed year.

```{r}
vars <- c(colnames(sv_spring_years[[1]]), colnames(sv_autumn_years[[1]]))

classify_var_location <- function(v) {
  if (str_detect(v, "rdr_")) return("Radar-level")
  if (str_detect(v, "fl_")) return("En-route")
  if (str_detect(v, "stp_")) return("Stopover")
  if (v %in% c("yday", "solar_elev", "solar_azim", "solar_elevation_change")) return("Phenology")
  if (v == "height") return("Height")
  if (v == "nr_birds") return("Radar-level")
}

classify_var_envir <- function(v) {
  if (str_detect(v, "_u|_v")) return("Wind")
  if (str_detect(v, "_t|_t2m")) return("Temperature")
  if (str_detect(v, "_lsrr|_crr")) return("Precipitation")
  if (str_detect(v, "_msl")) return("Pressure")
  if (str_detect(v, "_blh")) return("Boundary layer height")
  if (str_detect(v, "_cc|_lcc|_mcc|_hcc|_cbh")) return("Cloud cover")
  if (str_detect(v, "_r|_q")) return("Humidity")
  if (str_detect(v, "_lsm")) return("Land-sea mask")
  if (v == "height") return("Height")
  if (v == "yday") return("Day of year")
  if (str_detect(v, "solar_")) return("Time of night")
  if (str_detect(v, "_lat|_lon")) return("Take-off location")
  if (v == "nr_birds") return("Nr. of birds")
}

sv_list_collapsed_to_dataframe <- function(sv_list_collapsed, season, years) {
  sv_list_collapsed <- lapply(seq_along(sv_list_collapsed), function(i) {
    sv_listitem_collapsed_df <- as.data.frame(sv_list_collapsed[[i]])
    sv_listitem_collapsed_df$year_held_out <- years[[i]]
    sv_listitem_collapsed_df
  })
  sv_list_collapsed <- bind_rows(sv_list_collapsed)
  sv_list_collapsed$season <- season
  sv_list_collapsed
}

collapse_adjusted <- function(S, collapse_vars) {
  vars_present <- colnames(S)

  collapse_vars_filtered <- lapply(
    collapse_vars,
    function(x) x[x %in% vars_present]
  )

  collapse_shap(S, collapse_vars_filtered)
}

var_locations <- vapply(vars, classify_var_location, character(1))
var_envirs <- vapply(vars, classify_var_envir, character(1))

collapse_envirs <- split(vars, var_envirs)
collapse_envirs <- lapply(collapse_envirs, unique)

var_combo <- paste(var_envirs, var_locations, sep = "/")
collapse_combo <- split(vars, var_combo)
collapse_combo <- lapply(collapse_combo, unique)

sv_spring_years_envirs <- lapply(sv_spring_years, function(x) collapse_adjusted(x$S, collapse_envirs))
sv_autumn_years_envirs <- lapply(sv_autumn_years, function(x) collapse_adjusted(x$S, collapse_envirs))

sv_spring_years_combo <- lapply(sv_spring_years, function(x) collapse_adjusted(x$S, collapse_combo))
sv_autumn_years_combo <- lapply(sv_autumn_years, function(x) collapse_adjusted(x$S, collapse_combo))
spring_combo <- sv_list_collapsed_to_dataframe(sv_spring_years_combo, "spring", 2017:2024) %>%
  bind_cols(bind_rows(lapply(2017:2024, function(x) return(spring_ranks_and_datetimes))))

autumn_combo <- sv_list_collapsed_to_dataframe(sv_autumn_years_combo, "autumn", 2017:2024) %>%
  bind_cols(bind_rows(lapply(2017:2024, function(x) return(autumn_ranks_and_datetimes))))

spring_envirs <- sv_list_collapsed_to_dataframe(sv_spring_years_envirs, "spring", 2017:2024) %>%
  bind_cols(bind_rows(lapply(2017:2024, function(x) return(spring_ranks_and_datetimes))))

autumn_envirs <- sv_list_collapsed_to_dataframe(sv_autumn_years_envirs, "autumn", 2017:2024) %>%
  bind_cols(bind_rows(lapply(2017:2024, function(x) return(autumn_ranks_and_datetimes))))

all_combo <- bind_rows(spring_combo, autumn_combo)
all_envirs <- bind_rows(spring_envirs, autumn_envirs)

lookup_table <- data.frame(original = colnames(all_combo),
                           envir = unlist(lapply(colnames(all_combo), function(x) str_split(x, pattern = "/")[[1]][1])),
                           location = unlist(lapply(colnames(all_combo), function(x) str_split(x, pattern = "/")[[1]][2]))) %>%
  drop_na()

all_combo %>%
  data.table::as.data.table() %>%
  pivot_longer(cols = -c(year_held_out, season, period_rank_mtr, datetime, nightly_mt, nightly_mt_rank, mtr, year),
               names_to = "envir_combo", values_to = "shap") %>%
  identity() -> all_combo_preprocessed

all_envirs %>%
  data.table::as.data.table() %>%
  pivot_longer(cols = -c(year_held_out, season, period_rank_mtr, datetime, nightly_mt, nightly_mt_rank, mtr, year),
               names_to = "envir", values_to = "shap") %>%
  mutate(year = lubridate::year(datetime)) %>%
  identity() -> all_envirs_preprocessed

all_combo_preprocessed <- left_join(all_combo_preprocessed, lookup_table, by = c("envir_combo" = "original")) %>%
  mutate(year = lubridate::year(datetime)) %>%
  as.data.table()
```

## Nightly contributions by location

Contributions are averaged over observations within each night. Ranked panels align nights by migration traffic.

```{r}
DT <- as.data.table(all_combo_preprocessed)
DT <- DT[!envir %chin% c("Height", "Day of year", "Nr. of birds", "Time of night", "Land-sea mask")]
by_cols <- setdiff(names(DT), c("year_held_out", "shap"))
DT_avg <- DT[, .(shap = mean(shap, na.rm = TRUE)), by = by_cols]

DT_envirs <- as.data.table(all_envirs_preprocessed)
DT_envirs <- DT_envirs[!envir %chin% c("Height")]
by_cols <- setdiff(names(DT_envirs), c("year_held_out", "shap"))
DT_envirs_avg <- DT_envirs[, .(shap = mean(shap, na.rm = TRUE)), by = by_cols]

DT_avg %>%
  group_by(year, period_rank_mtr, envir, season, location, nightly_mt, nightly_mt_rank) %>%
  summarise(mean_shap = mean(shap),
            real_dt = min(datetime),
            disp_dt = update(real_dt, year = 2000)) %>%
  ungroup() %>%
  mutate(season_human = str_to_sentence(season),
         season_human = fct_relevel(as.factor(season_human), "Spring", "Autumn"),
         location = fct_relevel(as.factor(location), "Radar-level", "En-route", "Stopover"),
         envir = fct_relevel(as.factor(envir), "Wind", "Temperature", "Boundary layer height")) %>%
  identity() -> all_combo_small

DT_envirs_avg %>%
  group_by(year, period_rank_mtr, envir, season, nightly_mt, nightly_mt_rank) %>%
  summarise(mean_shap = mean(shap),
            real_dt = min(datetime),
            disp_dt = update(real_dt, year = 2000)) %>%
  ungroup() %>%
  mutate(season_human = str_to_sentence(season),
         season_human = fct_relevel(as.factor(season_human), "Spring", "Autumn")) %>%
  identity() -> all_envirs_small

all_combo_small %>%
  group_by(nightly_mt_rank, season, season_human) %>%
  summarise(nightly_mt = mean(nightly_mt),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_mt

all_combo_small %>%
  filter(envir %in% c("Wind", "Temperature", "Boundary layer height", "Precipitation")) %>%
  group_by(nightly_mt_rank, season, season_human, envir, location) %>%
  summarise(mean_shap = mean(mean_shap),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_shap

breaks_mt_ranked <- c("Avg.", "2017", "2018", "2019", "2020", "2021", "2022", "2023", "2024")
labels_mt_ranked <- c("Avg.", "", "2018", "", "2020", "", "2022", "", "2024")

all_combo_small %>%
  ungroup() %>%
  mutate(year_string = as.character(year)) %>%
  bind_rows(average_nightly_mt) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked)) %>%
  ggplot(aes(x = nightly_mt_rank, y = year_string, fill = nightly_mt)) +
  scale_fill_viridis_c(trans = cbrt_trans, oob = scales::squish, limit = c(0, 50000),
                       breaks = c(0, 10000, 25000, 50000), name = "Nightly migration traffic [birds/km]") +
  geom_tile() +
  scale_x_reverse(breaks = c(1, 10, 25, 50, 75, 100, 125)) +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  ggh4x::facet_grid2(cols = vars(season_human), strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    legend.position = "top",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "") -> plot_mt_ranked

all_combo_small %>%
  filter(envir %in% c("Wind", "Temperature", "Boundary layer height", "Precipitation")) %>%
  ungroup() %>%
  mutate(year_string = as.character(year),
         envir = forcats::fct_reorder(as.factor(envir), mean_shap, .fun = function(x) mean(abs(x), na.rm = TRUE), .desc = TRUE)) %>%
  bind_rows(average_nightly_shap) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked)) %>%
  ggplot(aes(x = nightly_mt_rank, y = year_string)) +
  geom_tile(aes(fill = mean_shap)) +
  scale_x_reverse(breaks = c(1, 10, 25, 50, 75, 100, 125)) +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  khroma::scale_fill_vik(name = "Mean SHAP value") +
  ggh4x::facet_grid2(rows = vars(envir, location), cols = vars(season_human), strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "Ranked night") -> plot_envir_ranked

plot_mt_ranked / plot_envir_ranked + plot_layout(heights = c(1, 8), axes = "collect")

ggsave("plots/mt_shap_ranked.png", width = 12, height = 14)
ggsave("plots/mt_shap_ranked.pdf", width = 12, height = 14)
```

In the chronological location panels, precipitation is shown per year but omitted from the average row.

```{r}
all_combo_small %>%
  mutate(disp_dt = as.Date(disp_dt)) %>%
  group_by(disp_dt, season, season_human) %>%
  summarise(nightly_mt = mean(nightly_mt),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_mt_chronological

all_combo_small %>%
  filter(envir %in% c("Wind", "Temperature", "Boundary layer height")) %>%
  mutate(disp_dt = as.Date(disp_dt)) %>%
  group_by(disp_dt, season, season_human, envir, location) %>%
  summarise(mean_shap = mean(mean_shap),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_shap_chronological


all_combo_small %>%
  ungroup() %>%
  mutate(year_string = as.character(year)) %>%
  bind_rows(average_nightly_mt_chronological) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked)) %>%
  ggplot(aes(x = as.Date(disp_dt), y = year_string, fill = nightly_mt)) +
  geom_tile() +
  scale_fill_viridis_c(trans = cbrt_trans, oob = scales::squish, limit = c(0, 50000),
                     breaks = c(0, 10000, 25000, 50000), name = "Nightly migration traffic [birds/km]") +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  facet_grid2(cols = vars(season_human), scales = "free_x", strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    legend.position = "top",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "") -> plot_mt_chronological

all_combo_small %>%
  filter(envir %in% c("Wind", "Temperature", "Boundary layer height", "Precipitation")) %>%
  ungroup() %>%
  mutate(year_string = as.character(year),
         envir = forcats::fct_reorder(as.factor(envir), mean_shap, .fun = function(x) mean(abs(x), na.rm = TRUE), .desc = TRUE)) %>%
  bind_rows(average_nightly_shap_chronological) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked)) %>%
  ggplot(aes(x = as.Date(disp_dt), y = year_string)) +
  geom_tile(aes(fill = mean_shap)) +
  scale_fill_vik(name = "Mean SHAP value") +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  facet_grid2(rows = vars(envir, location), cols = vars(season_human), scales = "free_x", strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "") -> plot_envir_chronological

plot_mt_chronological / plot_envir_chronological + plot_layout(heights = c(1, 8), axes = "collect")

ggsave("plots/mt_shap_chronological.png", width = 12, height = 14)
ggsave("plots/mt_shap_chronological.pdf", width = 12, height = 14)
```

## Nightly contributions by environment

These panels combine locations within each environmental group, in ranked and calendar order.

```{r}
all_envirs_small %>%
  group_by(nightly_mt_rank, season, season_human) %>%
  summarise(nightly_mt = mean(nightly_mt),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_mt_envirs

all_envirs_small %>%
  group_by(nightly_mt_rank, season, season_human, envir) %>%
  summarise(mean_shap = mean(mean_shap),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_shap_envirs

all_envirs_small %>%
  ungroup() %>%
  mutate(year_string = as.character(year)) %>%
  bind_rows(average_nightly_mt_envirs) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked)) %>%
  ggplot(aes(x = nightly_mt_rank, y = year_string, fill = nightly_mt)) +
  scale_fill_viridis_c(trans = cbrt_trans, oob = scales::squish, limit = c(0, 50000),
                       breaks = c(0, 10000, 25000, 50000), name = "Nightly migration traffic [birds/km]") +
  geom_tile() +
  scale_x_reverse(breaks = c(1, 10, 25, 50, 75, 100, 125)) +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  ggh4x::facet_grid2(cols = vars(season_human), strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    legend.position = "top",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "") -> plot_mt_ranked_envirs

all_envirs_small %>%
  ungroup() %>%
  mutate(year_string = as.character(year)) %>%
  bind_rows(average_nightly_shap_envirs) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked),
         envir_short = case_when(
           envir == "Boundary layer height" ~ "Blh.",
           envir == "Cloud cover" ~ "Cc.",
           envir == "Day of year" ~ "Day",
           envir == "Humidity" ~ "Hum.",
           envir == "Land-sea mask" ~ "Topog.",
           envir == "Nr. of birds" ~ "Sims.",
           envir == "Precipitation" ~ "Precip.",
           envir == "Pressure" ~ "Press.",
           envir == "Temperature" ~ "Temp.",
           envir == "Time of night" ~ "Time",
           envir == "Wind" ~ "Wind"
         ),
        envir_short = forcats::fct_reorder(as.factor(envir_short), mean_shap, .fun = function(x) mean(abs(x), na.rm = TRUE), .desc = TRUE)) %>%
  ggplot(aes(x = nightly_mt_rank, y = year_string)) +
  geom_tile(aes(fill = mean_shap)) +
  scale_x_reverse(breaks = c(1, 10, 25, 50, 75, 100, 125)) +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  khroma::scale_fill_vik(name = "Mean SHAP value") +
  ggh4x::facet_grid2(rows = vars(envir_short), cols = vars(season_human), strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc"),
  ) +
  labs(y = "Year", x = "Ranked night") -> plot_envir_ranked

plot_mt_ranked_envirs / plot_envir_ranked + plot_layout(heights = c(1, 11), axes = "collect")

ggsave("plots/mt_shap_envirs_ranked.png", width = 12, height = 14)
ggsave("plots/mt_shap_envirs_ranked.pdf", width = 12, height = 14)
```

```{r}
all_envirs_small %>%
  mutate(disp_dt = as.Date(disp_dt)) %>%
  group_by(disp_dt, season, season_human) %>%
  summarise(nightly_mt = mean(nightly_mt),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_mt_envirs_chronological

all_envirs_small %>%
  mutate(disp_dt = as.Date(disp_dt)) %>%
  group_by(disp_dt, season, season_human, envir) %>%
  summarise(mean_shap = mean(mean_shap),
            year_string = "Avg.") %>%
  ungroup() %>%
  identity() -> average_nightly_shap_envirs_chronological


all_envirs_small %>%
  ungroup() %>%
  mutate(year_string = as.character(year)) %>%
  bind_rows(average_nightly_mt_envirs_chronological) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked)) %>%
  ggplot(aes(x = as.Date(disp_dt), y = year_string, fill = nightly_mt)) +
  geom_tile() +
  scale_fill_viridis_c(trans = cbrt_trans, oob = scales::squish, limit = c(0, 50000),
                     breaks = c(0, 10000, 25000, 50000), name = "Nightly migration traffic [birds/km]") +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  facet_grid2(cols = vars(season_human), scales = "free_x", strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    legend.position = "top",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "") -> plot_mt_envirs_chronological

all_envirs_small %>%
  ungroup() %>%
  mutate(year_string = as.character(year)) %>%
  bind_rows(average_nightly_shap_envirs_chronological) %>%
  mutate(year_string = fct_relevel(as.factor(year_string), breaks_mt_ranked),
         envir_short = case_when(
           envir == "Boundary layer height" ~ "Blh.",
           envir == "Cloud cover" ~ "Cc.",
           envir == "Day of year" ~ "Day",
           envir == "Humidity" ~ "Hum.",
           envir == "Land-sea mask" ~ "Topog.",
           envir == "Nr. of birds" ~ "Sims.",
           envir == "Precipitation" ~ "Precip.",
           envir == "Pressure" ~ "Press.",
           envir == "Temperature" ~ "Temp.",
           envir == "Time of night" ~ "Time",
           envir == "Wind" ~ "Wind"
         ),
         envir_short = forcats::fct_reorder(as.factor(envir_short), mean_shap, .fun = function(x) mean(abs(x), na.rm = TRUE), .desc = TRUE)) %>%
  ggplot(aes(x = as.Date(disp_dt), y = year_string)) +
  geom_tile(aes(fill = mean_shap)) +
  scale_fill_vik(name = "Mean SHAP value") +
  scale_y_discrete(breaks = breaks_mt_ranked, labels = labels_mt_ranked) +
  facet_grid2(rows = vars(envir_short), cols = vars(season_human), scales = "free_x", strip = strip_nested(bleed = FALSE)) +
  coord_cartesian(expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.justification = "right",
    legend.box.just = "right",
    legend.key.width = unit(0.08, "npc")
  ) +
  labs(y = "Year", x = "") -> plot_envir_envirs_chronological

plot_mt_envirs_chronological / plot_envir_envirs_chronological + plot_layout(heights = c(1, 11), axes = "collect")

ggsave("plots/mt_shap_envirs_chronological.png", width = 12, height = 14)
ggsave("plots/mt_shap_envirs_chronological.pdf", width = 12, height = 14)
```
